POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2023-10-20 technical-report/main.tex
Move theorem definitions out of macros for Springer compatibility.
adding dual presentation of session types (with intuitionistic and linear context separated). Now this is called in technical-report but it can be enabled/disabled elsewhere (file appst.tex)
Split challenges.tex into multiple files.
removed appst appendix from tech report
Fix compilation error in technical report.
Backport Alceste's changes into challenges/.
update technical report

\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{draftwatermark}

\SetWatermarkText{Draft}
\SetWatermarkScale{5}
\SetWatermarkColor[gray]{0.90}

@@

\title{POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types}

@@

In recent years, the programming language community has embraced
machine checked proofs for theoretical results in papers. The
POPLMark~\cite{POPLMark} challenge was proposed for this purpose,
and to get the community on track. Since then the number of formal
proofs accompanying papers has grown\footnote{add some POPL/PLDI/etc
  statistics here}. However, formal proofs are not equally spread in
the field. In this call to action, we propose to extend the key
contribution of challenges like POPLMark is to help develop the
necessary momentum and culture to be able to formalise results to the
concurrent calculi community by proposing a challenge the exercises
the key aspects of reasoning about process calculi. We propose this
document, and a \href{https://concurrentbenchmark.github.io/}{website}
to enable the community to collaborate on the project.

It has been a long time since the original challenge, and many changes
have happened. Particularly, POPLMark and any challenge cannot address
every conceivable technique. Challenges have to be simple so that the
techniques are not shadowed by technical details and also feasible to
implement without heroic efforts. In that vein, it is apparent that
new challenges are needed, work like POPLMark
Reloaded~\cite{POPLMarkReloaded} strives to extend the scope of the
original challenge to proofs using logical relations. This technique
is crucial to many current developments.

As mentioned before, in this work we address another expansion of
scope for such challenges, going from the $\lambda$-calculus to
process calculi such as the $\pi$-calculus and session type systems
(and other behavioural type systems). These systems require particular
insights to implement. So far, we have identified three key
aspects\footnote{We want to keep the number low, but the exact nature
  of the challenges is still open to discussion (as are most things in
  this project).}. First, the idea of scope extrusion in name passing
calculi. This generalises the idea of binders in ways that techniques
that were adequate for $\lambda$-calculus binders are not so
convenient for name passing systems. Second, session types in
particular, and behavioural types in general often depend on linearity
in their type systems. The formalisation of linear type systems posses
a challenge given that it is not uniformly well supported among the
many proof assistants. Finally, processes in concurrent systems often
contain infinite behaviour that is typically modelled with
coinduction. This technique is used to reason about (bi)similarity
between processes, the equivalence of their traces, or subtyping
between infinite trees of possible behaviours. This challenge
concentrates on those issues.

In this work we propose a challenge where each part independently
exercises one of the three areas, however it is often the case that a
problem requires more than one of these techniques simultaneously. In
this work, we explore them in isolation, in order to facilitate the
implementation of smaller solutions. Concretely, we do not address the
idea of how to combine these techniques, leaving the discussion of
that to the future when we have collected enough entries to these
problems.

We do not claim in this work that these problems have never been
approached and formalised before. In fact, the contrary is true. All
of these problems have been addressed in several forms. This motivates
this challenge to enable an easier comparison between solutions (since
they all implement the same challenge), and in a setting that is
\emph{simple enough to not be distracting, yet complex enough that we can
compare strengths and weaknesses of each}.

The rest of this paper is structured in the following way:
\cref{sec:rationale} expands on the rationale for the selected
challenge problems. Then \cref{sec:problems} presents the problems. We
discuss the solutions in \cref{sec:solutions}, in particular
\cref{sec:tutorial} walks through a tutorial solution. And finally
\cref{sec:conclusion} presents some conclusions and it contains a call
to action to invite the submission of new solutions.

\section{Rationale for choosing the problems}\label{sec:rationale}

 The idea is to explore three areas: scope
extrusion in name passing calculi, linearity requirements in
behavioural types, and finally coinductive reasoning in trace
equivalence, (bi)similarity, or subtyping in the presence of
recursion.

The idea is to have three small problems in the aforementioned
subjects that are independent of each other and that are easy to
understand and to implement. This is important so we have more
entries. They have to be small so it is not a lot to read, and
independent so people may choose to implement only one of them, or
work on them in any order. Moreover, the specification should only
need to be met in spirit. The idea is to provide a solution author
with some space to show the best aspects of their solution even if
that means adapting the problem definition while keeping the spirit of
the challenge.

Finally, it will be important to mention in the rationale for problems that
the key aspects are:
\begin{itemize}
\item Scope extrusion
\item Linearity
\item Coinduction
\end{itemize}

But the exact presentation of the concrete problems is somewhat
flexible. In the sense that the use of certain techniques or tools may
require changing the presentation of the system to fit the technique.
We are more interested in the techniques that are applicable to these
problems as a class of problems than to the exact specification we
propose. Of course, for clarity if a solution changes the definition
it should document in what sense the definition differs. So changes
allowed, and even encouraged, but the result should be ``morally'' the
same system and there should be a rationale for the changes in
presentation. That way we not only see valid techniques but also what
presentation of the rules to use and the rationale to chose a
particular approach.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\textwidth]{../images/benchmark.jpeg}
  \caption{The benchmark's big picture.}
  \label{fig:bigpic}
\end{figure}

The big picture for the benchmark can be seen in \cref{fig:bigpic}
where the three pillars (i.e.: concepts typically needed for process
calculus proofs) appear in blue, with some ideas of what to prove. And
a second step of combining the three aspects (for a total of six steps
for the whole benchmark), and of course the $+1$-step for the
combination of everything on the reader's next research papers.

\input{../challenges.tex}

\section{The solutions}\label{sec:solutions}

\subsection{A tutorial solution on paper}\label{sec:tutorial}

Probably the skeleton here and more details on an appendix.

\subsection{Flexibility in solving a challenge}

\dots

\section{Future work}
\begin{itemize}
\item Multiparty session types
\item Choreography
\item Encodings
\item Code extraction
\item Conversation types
\item Psi-calculus
\item Model checking for system properties
\item More general types than session types
\end{itemize}

\section{Conclusion and related work} \label{sec:conclusion}

>>>>

\documentclass{article}
\synctex=1
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{mathpartir}
\usepackage{cite}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{hyperref}
\usepackage[inline]{enumitem}
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

\usepackage[status=draft]{fixme}
\fxsetup{theme=color}

%\SetWatermarkText{Draft}
%\SetWatermarkScale{5}
%\SetWatermarkColor[gray]{0.90}

@@

\title{POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types}
\author{...}

@@

% The abstract of the influential POPLMark challenge~\cite{POPLMark}
% starts with an important question: \emph{``How close are we to a world
%   where every paper on programming languages is accompanied by an
%   electronic appendix with machine-checked proofs?''} The introduction
% of the POPLMark challenge spearheaded a shift towards publications
% that include mechanised proofs, fostering the progress of proof
% assistants, libraries, and best practices. Later on, the follow-up
% POPLMark Reloaded challenge~\cite{POPLMarkReloaded} has encouraged the
% development of best practices and tool support for proofs using
% logical relations.

The authors of both POPLMark and POPLMark Reloaded explicitly
note that their benchmarks were only a beginning.  Furthermore, they
specifically mention reasoning about concurrency using coinduction and
linear environments as points for future work.  In this spirit, in order to
further increase the reach of the POPLMark challenges, we propose a new
collection of benchmark problems specifically designed to address typical issues
that arise when mechanising message-passing concurrency using process calculi.

Several papers in the realm of process calculi already include mechanised proofs
(e.g.~\cite{DBLP:conf/pldi/Castro-Perez0GY21,DBLP:conf/tacas/CastroFY20,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or
discuss proof mechanisation techniques
(e.g.~\cite{DBLP:journals/jar/BengtsonPW16, DBLP:conf/tphol/Gay01,
  DBLP:conf/ppdp/Thiemann19, DBLP:conf/forte/ZalakainD21}).  However, our
experience is that choosing the appropriate proof mechanisation techniques and
tools remains a significant challenge with few guidelines.  This often leads
researchers towards trial-and-error, attempting sub-optimal and ad-hoc
solutions; this increases the overall effort required for proof mechanisation,
and hampers the reuse of techniques and results.
%
Therefore, it is our opinion that the research community in message-passing
concurrency (and more specifically, in process calculi) will benefit from a set
of solutions exploring different ways of mechanising typical problems in the
field.
Like the authors of the POPLMark challenges, we seek to
answer a number of questions:
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} What techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} What improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}
We have identified three key aspects which typically cause difficulties when
mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion},
and \emph{coinductive reasoning} for infinite behaviours.
Our benchmark problems are designed to focus on these aspects (discussed
in more detail in \cref{sec:challenge-problems}), with three main goals:

\begin{enumerate}[label=\textbf{(G\arabic*)},leftmargin=10mm]
\item\label{item:goal-comperison-accessibility} Enable the comparison of
  different proof mechanisation approaches, making the challenges accessible to
  mechanisation experts who may not be familiar with concurrency theory.

\item\label{item:goal-tutorials} Encourage the development of guidelines and
  tutorials demonstrating and comparing the many available proof mechanisation
  techniques, libraries, and proof assistant features.

\item\label{item:goal-reusability} Prioritise the exploration of mechanisation
  techniques that may be useful and reusable for future research.
\end{enumerate}
We also aim at strengthening the culture of mechanisation, by rallying the
community to collaborate on exploring and developing new tools and techniques.

The three aspects we selected (linearity, scope extrusion, coinduction) are of
course not the only ones that may cause issues in mechanisations --- but they
are fundamental to concurrency theory, and they emerge in most mechanisations.
The mechanisation of novel research results often requires addressing more than
just one of these aspects at once --- and we see the combination of techniques
as a next step beyond the scope of this challenge.%  (we discuss this in
% \cref{sec:going-beyond}).

% Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
% While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the longer term, we expect to use the website for promoting best practices,
tutorials and guidelines derived from solutions to our challenges.
We encourage anyone interested to try the challenges using their
favourite tools and techniques, and to send us their solutions.

@@

\section{The challenges}\label{sec:challenge-problems}
\label{sec:prelim}
First, we list some common notions and conventions that we use in the
challenges. Since the calculi under study are somewhat different, each
section lists the changes that apply. The challenges are inspired by
simplified versions of calculi (and proofs) from chapters 1 and 2
of~\cite{picalcbook} and~\cite{Vasconcelos2012} where the idea is to
use sources that can be considered standard and close to current practice.

We assume the existence of some set of \emph{base values}, represented by the symbols
\( a, b, \dots \), the existence of some set
of \emph{variables}, represented by the symbols
\( l, m, \dots \), and the existence of some set of \emph{names},
represented by the symbols \( x, y, \dots \).%
\footnote{%
  Unlike the standard $\pi$-calculus, we distinguish variables from names to
  better control the expressiveness of the calculi under study, and the scope of
  the corresponding challenges: the key distinction is that names are used as communication channels (and can be sent and received in the scope extrusion challenge), whereas variables are only bound by inputs and cannot be restricted, sent nor received, in the style of value-passing CCS \cite{Milner1989}.%
} %
We assume that all of these sets are infinite and that their elements can be
compared for equality.

The syntax of processes includes:
the process \( \Pend \) or \emph{inaction}, a process which can do nothing. The process \( \Ppar{P}{Q} \) is the \emph{parallel composition} of process \( P \) and process \( Q \).
The two components can proceed independently of each other, or they can interact via shared names.
%The process \( \Pchoice{P}{Q} \) is a non-deterministic \emph{choice} between continuing as the process \( P \) or as the process \( Q \).

For communication, processes include \emph{input} and \emph{output}, whose
signature depends on the calculus being value-passing or name-passing.
We use here the metavariables $c,k$ to abstract over this choice
--- i.e.~$c$ may be either a value or a variable or a name, whereas $k$ may
be a variable or a name.
The process \( \Pout{x}{c}{P} \) is an \emph{output}, which can send
\( c \) via \( x \), then continue as \( P \).  % The intention is that
% the value \( v \) must be a base value when it is actually sent, and
% this will be enforced in the semantics later on
The process \( \Pin{x}{k}{P} \) is an \emph{input}, which can receive a $c$
via \( x \), then continue as \( P \) with the received element
substituted for \( k \).  The input operator thus
binds \( k \) in \( P \).

The process \( \Pres{x}{P} \) is the \emph{restriction} of the name
\( x \) to \( P \), binding \( x \) in \( P \).

The process \( \Preplicate{P} \) is the \emph{replication} of the process \( P \).
It can be thought of as the infinite composition \( \Ppar{P}{\Ppar{P}{\cdots}} \).
Replication makes it possible to express infinite behaviours.

We use the notation \( \freenames{P} \) to denote the set of
names that occur free,
% (i.e.\ not bound by a restriction) in \( P \),
\( \boundnames{P} \) to denote the set of names that occur bound
% (by a restriction)
in \( P \) and  \( \freevars{P} \)
to denote the set of variables that occur free
% (i.e.\ not bound by an input)
in \( P \).  We use the notation \( \boundvars{P} \)
for the set of variables that occur bound
% (by an input)
in \( P \).  We use the notation \( \subst{P}{a}{l} \) to denote
the process \( P \) with base value \( a \) substituted for variable
\( l \). Similarly, \( \subst{P}{x}{y} \) denotes the process
\( P \) with name \( x \) substituted for name \( y \).
We use the notation \( \applysubst{\sigma}{P} \) to denote the process \( P \) with a finite number of arbitrary substitutions applied to it.

Two processes \( P \) and \( Q \) are \( \alpha \)-convertible,
written \( \alphacon{P}{Q} \), if \( Q \) can be obtained from \( P \)
by a finite number of substitutions of bound variables.  As a
convention, we identify \( \alpha \)-convertible processes and we
assume that bound names and bound variables of any processes are
chosen to be different from the names and variables that occur
free in any other entities under consideration, such as processes,
substitutions, and sets of names or variables.  This is justified
because any overlapping names and variables may be
\( \alpha \)-converted such that the assumption is satisfied.

A \emph{context} is obtained by taking a process and replacing a single occurrence of \( \Pend \) in it with the special \emph{hole} symbol \( \ctxhole \).
As a convention, we do \emph{not} identify \( \alpha \)-convertible contexts.
%
A context acts as a function between processes:
a context \( C \) can be \emph{applied} to a process \( P \), written \( \applyctx{C}{P} \), by replacing the hole in C by \( P \), thus obtaining another process.
The replacement is literal, so names and variables that are free in \( P \) can become bound in \( \applyctx{C}{P} \).

We say that an equivalence relation \( \mathcal{S} \) is a \emph{congruence} if \( (P,Q) \in \mathcal{S} \) implies that for any context \( C \), \( (\applyctx{C}{P}, \applyctx{C}{Q}) \in \mathcal{S} \).

\subsection{Challenge: Linearity and behavioral type systems}
\label{sec:challenge:linearity-beh-types}
This challenge formalises a proof that requires reasoning about linearity of channels.
Linearity is the notion that a channel must be used exactly once in a process.
This is necessary to prove properties about session type systems, and the key
issue of this challenge is reasoning about the linearity of context splitting operations.
Linear reasoning is also necessary to formalise e.g.\ linear and affine types for the pi-calculus and cut elimination in linear logics.

The setting for this challenge is a small calculus with a session type
system, the syntax and semantics of which are given below. The
calculus is a fragment of the one presented in~\cite{Vasconcelos2012},
formulated in the dual style of~\cite{barber96tr}.

The main objective of this challenge is to prove type preservation (also
known as subject reduction), i.e.\ that well-typed
processes can only transition to processes which are also well-typed
in the same context.
The second objective is to prove type safety, i.e.\ that well-typed processes are also well-formed in the sense that they do not use endpoints in a non-dual way.

\subsubsection{Syntax.}
The syntax is given by the grammar
\[
\begin{array}{r@{\qquad}c@{\qquad}l}
  v,w & \Coloneqq & a \quad\mid\quad l \\
   P,Q & \Coloneqq & \Pend \quad\mid\quad \Pout{x}{v}{P} \quad\mid\quad \Pin{x}{l}{P} \quad\mid\quad \PBpar{P}{Q} \quad\mid\quad  \Presd{x}{y}{P}
\end{array}
\]
where a \emph{value} \( v, w, \dots \) is either a base value \( a \) or a variable \( l \).

The output process \( \Pout{x}{v}{P} \) sends the value \( v \) via \( x \) and then continues as \( P \).
The intention is that the value \( v \) must be a base value when it is actually sent, and this is enforced in the semantics later on.
The input process \( \Pin{x}{l}{P} \) waits for a base value from \( x \) and then continues as \( P \) with the received value substituted for the variable \( l \).
%
The process \( \Presd{x}{y}{P} \) represents a \emph{session} with endpoints named \( x \) and \( y \) which are bound in \( P \). In \( P \), the names \( x \) and \( y \) can be used to exchange messages over the session (sending on \( x \) and receiving on \( y \) or vice versa).
%
%Finally, the process $\PBpar{P}{Q}$ represents the parallel composition of $P$ and $Q$.
%
Note that in this calculus channels cannot be sent in messages, therefore the topology of the communication network described by a process cannot change.
Also, there is no recursion or replication in the syntax, hence no infinite behaviors can be expressed. In particular, we only model linear (as opposed to shared) channels.

\subsubsection{Semantics.}
We describe the actions that the system can perform through a small step operational semantics.
As usual, we use  a \emph{structural congruence} relation that equates processes that we deem to be indistinguishable. Structural congruence is the smallest congruence relation that satisfies the following axioms:
\begin{mathpar}
  \inferrule[Sc-Par-Comm]{ }{\scong{\Ppar{P}{Q}}{\Ppar{Q}{P}}}
  \and
  \inferrule[Sc-Par-Assoc]{ }{\scong{\Ppar{(\Ppar{P}{Q})}{R}}{\Ppar{P}{(\Ppar{Q}{R})}}}
  \and
  \inferrule[Sc-Par-Inact]{ }{\scong{\Ppar{P}{\Pend}}{P}}
  \and
  \inferrule[Sc-Res-Par]{\{x,y\} \cap \freenames{Q} = \emptyset}{\scong{\Ppar{\Presd{x}{y}{P}}{Q}}{\Presd{x}{y}{(\Ppar{P}{Q})}}}
  \and
  \inferrule[Sc-Res-Inact]{ }{\scong{\Presd{x}{y}{\Pend}}{\Pend}}
  \and
  \inferrule[Sc-Res]{ }{\scong{\Presd{x_1}{y_1}{\Presd{x_2}{y_2}{P}}}{\Presd{x_2}{y_2}{\Presd{x_1}{y_1}{P}}}}
\end{mathpar}

The operational semantics is defined as the following relation on processes:
\begin{mathparpagebreakable}
  \inferrule[R-Com]{ }{\reduces{\Presd{x}{y}{(\Ppar{\Pout{x}{a}{P}}{\Ppar{\Pin{y}{l}{Q}}{R}})}}{\Presd{x}{y}{(\Ppar{P}{\Ppar{\subst{Q}{a}{l}}{R}})}}}
  \and
  \inferrule[R-Res]{\reduces{P}{Q}}{\reduces{\Presd{x}{y}{P}}{\Presd{x}{y}{Q}}}
  \and
  \inferrule[R-Par]{\reduces{P}{Q}}{\reduces{\Ppar{P}{R}}{\Ppar{Q}{R}}}
  \and
  \inferrule[R-Struct]{\scong{P}{P'} \\ \reduces{P'}{Q'} \\ \scong{Q}{Q'}}{\reduces{P}{Q}}
\end{mathparpagebreakable}

Note that reductions are allowed only for restricted pairs of session endpoints. This makes it possible to formulate subject reduction so that the typing context is exactly the same before and after each reduction.
%
Note also that due to rule \TirName{R-Com}, the process \( \Pin{y}{l}{P} \) can receive \emph{any} base value.
Since the rule \TirName{R-Com} only applies to sending base values, there is no way to send a variable or a name.

\subsubsection{Session types.}
Our process syntax allows us to write processes that are ill formed in
the sense that they either use the endpoints bound by a restriction to
communicate in a way that does not follow the intended duality, or
attempt to send something which is not a base value.  As an example,
the process
\( \Presd{x}{y}{(\Ppar{\Pout{x}{a}{\Pend}}{\Pout{y}{a}{\Pend}})} \)
attempts to send a base value on both \( x \) and \( y\), whereas one
of the names should be used for receiving in order to guarantee
progress.  Another example is the process
\( \Presd{x}{y}{(\Ppar{\Pout{x}{l}{\Pend}}{\Pin{y}{l}{\Pend}})} \),
which attempts to send a variable that is not instantiated at the time
of sending.
%
To prevent these issues, we introduce a \emph{session type system} which
rules out ill-formed processes.

\subsubsection{Syntax.}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \), unrestricted type contexts \( \Gamma \) and linear type contexts \( \Delta \) is as follows:
\[
  \begin{array}{r@{\qquad}c@{\qquad}l}
  S,T & \Coloneqq & \Tend \quad\mid\quad \Tbase \quad\mid\quad \Tin{S} \quad\mid\quad \Tout{S} \\
    \Gamma & \Coloneqq & \Cempty \quad\mid\quad \Gamma, l \\
                    \Delta &::= & \Cempty \quad\mid\quad \Cadd{\Delta}{\hastype{x}{S}}
  \end{array}
\]
The \emph{end type} \( \Tend \) describes an endpoint on which no further interaction is possible.
The \emph{base type} \( \Tbase \) describes base values.
The \emph{input type} \( \Tin{S} \) describes endpoints used for receiving a value and then according to \( S \).
The \emph{output type} \( \Tout{S} \) describes endpoints used for sending a value and then according to \( S \).

Typing contexts gather type information about names and variables.
\emph{Unrestricted} contexts are simply sets of names since we only have one
base type. \emph{Linear contexts} associate a type to endpoints. We use
the comma as split/union, overloaded to singletons, and \( \Cempty \) as the
empty context. We extend the Barendregt convention to contexts, so that all
entries are distinct.  Note that the order in which information is added to a
type context does not matter.

Since we need to determine whether endpoints are used in complementary ways to determine whether processes are well formed, we need to formally define the dual of a type as follows:
\begin{mathpar}
  \inferrule{}{\dual{\Tin{S}} = \Tout{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tout{S}} = \Tin{\dual{S}}}
  \and
  \inferrule{}{\dual{\Tend} = \Tend}
\end{mathpar}
Note that the dual function is partial since it is undefined for the base type.

\subsubsection{Typing rules.}
Our type system is aimed at maintaining two invariants:
\begin{enumerate}
\item No endpoint is used simultaneously by parallel processes;
\item The two endpoints of the same session have dual types.
\end{enumerate}
The first invariant is maintained by linearly splitting type contexts when typing compositions of processes, the second by requiring duality when typing restrictions.

We have two typing judgments: one for values, and one for processes.
The typing rules for values are:
\begin{mathpar}
  \inferrule[T-Base]{ }{\typev{\Gamma}{\hastype{a}{\Tbase}}} \and
  \inferrule[T-Var]{ }{\typev{\Cadd{\Gamma}l }{\hastype{l}{\Tbase}}}
\end{mathpar}
% In the typing rule for names, $\tend \Phi$ denotes that all types in the range of $\Phi$ are $\Tend$:
% \begin{mathpar}
%   \inferrule[T-Name]{{\tend\Delta }}{\types{{\Gamma; (\Delta,\hastype{x}{T})}}{\hastype{x}{T}}}
% \end{mathpar}
The typing rules for processes are as follows:
\begin{mathpar}

  \inferrule[T-Inact]{\tend\Delta }{\types{\Gamma;\Delta}{\Pend}}
  \and
  \inferrule[T-Par]{\types{\Gamma;\Delta_1}{P} \\ \types{\Gamma;\Delta_2}{Q}}
  {\types{\Gamma; \Csplit{\Delta_1}{\Delta_2}}{\Ppar{P}{Q}}}
  \and
  \inferrule[T-Res]{\types{\Gamma; (\Cadd{\Cadd{\Delta}{\hastype{x}{T}}}{\hastype{y}{\dual{T}}}}{P})}{\types{\Gamma}{\Presd{x}{y}{P}}}
  \and
    \inferrule[T-Out]{
      \typev{\Gamma}{\hastype{v}{\Tbase}} \\ \types{\Gamma; \Cupdate{\Delta}{\hastype{x}{T}}}{P}}{\types{\Gamma; (\Csplit{\Delta}{\hastype{x}{\Tout{T}}})}{\Pout{x}{v}{P}}}
    \and
    \inferrule[T-IN]{
      \types{(\Gamma,  l ); (\Cupdate{\Delta}{\hastype{x}{T}})}{P}}{\types{\Gamma; (\Csplit{\Delta}{\hastype{x}{\Tin{T}}})}{\Pin{x}{l}{P}}}
\end{mathpar}
Note that we do not need a judgment for typing channels, since it is already folded into the \TirName{T-In} and \TirName{T-Out} rules.

\subsubsection{Challenge.}
The objective of this challenge is to prove subject reduction and type safety for our calculus with session types. We start with:
\begin{theorem}[Subject reduction]
  \label{thm:subject-reduction}
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}

The following lemmas are needed:
\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \) then
    \( \types{(\Cadd{\Gamma}{l});\Delta}{P} \).
  \item If \( \types{\Gamma; \Delta}{P} \) % and \( \tend{\Delta} \)
    then \( \types{\Gamma;(\Cadd{\Delta}{\hastype{x}{\Tend}})}{P} \).
  \end{enumerate}
\end{lemma}

\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \( \types{\Gamma; (\Cadd{\Delta}{\hastype{x}{T}})}{P} \) and
  $x\not\in \freenames{P}$, then \( \types{\Gamma; \Delta}{P} \).
\end{lemma}

\begin{lemma}[Substitution]\mbox{}
  \label{le:subst}
  If $\types{(\Cadd{\Gamma}{l});\Delta}{P}$ and
  ${\typev{\Gamma}{\hastype{a}{\Tbase}}}$ then
  \( \types{\Gamma;\Delta}{\subst{P}{a}{l}} \).
\end{lemma}

To prove that congruence preserves typing we need to spell out in more detail the former. First, we denote with $\sconga\cdot\cdot$ the relation induced by the six axioms.

\begin{lemma}[Preservation for $\stackrel{a}{\equiv}$]
  \label{le:presequiva}
  If \( \sconga{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}

Now we give rules for the compatible
equivalence relation induced by $\sconga\cdot\cdot$, which we still
write as $\scong\cdot\cdot$: namely, add to reflexivity, symmetry and
transitivity the following condition:
 \begin{mathpar}
   \inferrule[Cong]{\sconga P Q}{\scong {C[P]}{C[Q]}}
 \end{mathpar}

\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
    If \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}

To formulate safety, we need to formally define what we mean by
well-formed process.  We say that a process \( P \) is \emph{prefixed
  at variable \( x \)} if \( P\equiv\Pout{x}{v}{P'} \) or
\( P\equiv\Pin{x}{l}{P'} \) for some $P'$.
%
A process $P$ is then \emph{well formed} if, for every $P_1$, $P_2$,
and $R$ such that
\( P\equiv \Presd{x_1}{y_1}{\dots
  \Presd{x_n}{y_n}{(\Ppar{\Ppar{P_1}{P_2}}{R})}} \), with
\( n \geq 0 \), it holds that, if \( P_1 \) is prefixed at \( x_1 \)
and \( P_2 \) is prefixed at \( y_1 \) (or vice versa), then
\( \Ppar{P_1}{P_2}
\equiv\Ppar{\Pout{x_1}{a}{P_1'}}{\Pin{y_1}{l}{P_2'}} \), for some
$P_1'$ and $P_2'$.

Note that well-formed processes do not necessarily reduce. For example, the process
\begin{equation*}
  \Presd{x_1}{y_1}{\Presd{x_2}{y_2}{(\Ppar{\Pout{x_1}{a}{\Pin{y_2}{l}{\Pend}}}{\Pout{y_2}{x_2}{\Pin{y_1}{l}{\Pend}}})}}
\end{equation*}
is well formed but also irreducible.

\begin{theorem}[Type safety]
  \label{thm:type-safety}
  If \( \types{\Cempty}{P} \), then \( P \) is well formed.
\end{theorem}

\begin{corollary}
  If \( \types{\Cempty}{P} \) and \( \reduces{P}{Q} \) then \( Q \) is well formed.
\end{corollary}

\subsection{Challenge: Name passing and scope extrusion}
\label{sec:challenge:name-passing-scope-extrusion}

This challenge formalises a proof that requires explicit scope extrusion.
Scope extrusion is the notion that a process can send restricted names to another process, as long as the restriction can safely be ``extruded'' (i.e.\ expanded) to include the receiving process.
This e.g.\ allows a process to set up a private connection by sending a restricted name to another process, then using this name for further communication.
The key issue of this challenge is reasoning about names that are ``in the process'' of being scope-extruded, which often presents difficulties for the mechanisation of binders.

Reasoning about scope extrusion explicitly can sometimes be avoided by introducing a structural congruence rule into the semantics, but doing this means we lose information about the scope when reasoning about the semantics.
Explicitly reasoning about scope extrusion is necessary to describe e.g.\ runtime monitors and compositions of systems.

The setting for this challenge is a ``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in the other challenges) names can be sent and received, and bound by input constructs (similarly to variables in the other calculi).
We define two different semantics for our system: one that avoids explicit reasoning about scope extrusion, and one that does not.
The objective of this challenge is to prove that the two semantics are equivalent up to structural congruence.

\subsubsection{Syntax.}
The syntax of processes is given by:
\[
\begin{array}{r@{\qquad}c@{\qquad}l}
  P,Q & \Coloneqq & \Pend \quad\mid\quad \PBpar{P}{Q} \quad\mid\quad \Pout{x}{y}{P} \quad\mid\quad \Pin{x}{y}{P} \quad\mid\quad \Pres{x}{P}
\end{array}
\]

%The process \( \Pend \) is \emph{inaction}: a process which can do nothing.
The process \( \Pout{x}{y}{P} \) is an \emph{output}, which can send the name \( y \) via \( x \), then continue as \( P \).
The process \( \Pin{x}{y}{P} \) is an \emph{input}, which can receive a name via \( x \), then continue as \( P \) with the received name substituted for \( y \).
The input operator thus binds the name \( y \) in \( P \).
% The process \( \Ppar{P}{Q} \) is the \emph{composition} of process \( P \) and process \( Q \).
% The two components can proceed independently of each other, or they can interact via shared names.
% The process \( \Pres{x}{P} \) is the \emph{restriction} of the name \( x \) to \( P \).
% Components in \( P \) can use the name \( x \) to interact with each other, but not with processes outside of the restriction.
% The restriction operator thus binds the name \( x \) in \( P \).
Note that the scope of a restriction may change when processes interact. Namely, a restricted name may be sent \emph{outside} of its scope.
% The process \( \Pchoice{P}{Q} \) is a non-deterministic \emph{choice} between continuing as the process \( P \) or as the process \( Q \).
Note that there is no recursion or replication in the syntax, and thus no infinite behaviors can be expressed.
This simplifies the theory and is orthogonal to the concept of scope extrusion.

% We will use the notation \( \freenames{P} \) to denote the set of names that occur free (i.e.\ not bound by a restriction or an input) in \( P \).
% We will use the notation \( \boundnames{P} \) to denote the set of names that occur bound (by a restriction or an input) in \( P \).
% We will use the notation \( \subst{P}{x}{y} \) to denote the process \( P \) with \( x \) substituted for \( y \).

% Two processes \( P \) and \( Q \) are \( \alpha \)-convertible, written \( \alphacon{P}{Q} \), if \( Q \) can be obtained from \( P \) by a finite number of substitutions of bound names.
% As a convention, we will identify \( \alpha \)-convertible processes.

% As a convention, we assume that the bound names occurring in any collection of processes are chosen to be different from the free names occurring in those processes and from the names occurring in any substitutions applied to the processes.
% This is justified because any overlapping names may be \( \alpha \)-converted such that the assumption is satisfied.

\subsubsection{Reduction semantics.}
The first semantics is an operational reduction semantics, which avoids reasoning explicitly about scope extrusion by way of a structural congruence rule.
\emph{Structural congruence} is the smallest congruence relation that satisfies the following axioms:

\begin{mathpar}
  \inferrule[Sc-Par-Assoc]{ }{\scong{\Ppar{P}{(\Ppar{Q}{R})}}{\Ppar{(\Ppar{P}{Q})}{R}}}
  \and
  \inferrule[Sc-Par-Comm]{ }{\scong{\Ppar{P}{Q}}{\Ppar{Q}{P}}}
  \and
  \inferrule[Sc-Par-Inact]{ }{\scong{\Ppar{P}{\Pend}}{P}}
  \and
  \inferrule[Sc-Res-Par]{x \notin \freenames{Q}}{\scong{\Ppar{\Pres{x}{P}}{Q}}{\Pres{x}{(\Ppar{P}{Q})}}}
  \and
  \inferrule[Sc-Res-Inact]{ }{\scong{\Pres{x}{\Pend}}{\Pend}}
  \and
  \inferrule[Sc-Res]{ }{\scong{\Pres{x}{\Pres{y}{P}}}{\Pres{y}{\Pres{x}{P}}}}
\end{mathpar}
